Towards Detecting Stalkerware Through Network Traces

### 2019-03-18 paper/main.tex
Edited introduction

\section{Introduction}
In a world where technology now pervades every aspect of our lives, some 
aspects of our relationships with intimate partners have changed too. 
Specifically, when these relationships become abusive, there are more options 
available to the abuser for attempting to retain control over their partner. 
One of these options is the ability to install monitoring software on their 
partner's phone or computer to record their online activity, communications, 
and location. We refer to such software as ``stalkerware.''  In a survey of 72 
domestic violence shelters performed by NPR in 2014, 85\% said they worked with 
survivors whose abusers had tracked their location using their smartphone's 
GPS. 75\% said they are working with clients who have had their conversations 
monitored by hidden stalkerware apps.\cite{shahani_smartphones}

Although stalkerware is superficially similar to other types of malware, a few 
key differences exist between them. First, stalkerware is meant to be installed 
deliberately, by someone with physical access to the device that is to be 
monitored. In the context of intimate partner abuse, this presents an enormous 
threat. An abuser will often know or be able to coerce passwords to the device 
from the victim, making traditional defense measures irrelevant. Second, the 
threat model used by malware developers includes sophisticated antivirus and 
intrusion detection systems (IDSs). Malware is therefore much more likely to 
attempt to obfuscate its network traffic. Third, stalkerware falls for the most 
part into two categories: apps that are explicitly advertised for surveilling 
spouses and partners, and apps that have a different stated use but can be 
repurposed to stalk a target. These are usually marketed for theft prevention, 
parental control of children, and employer control of employees. Previous 
literature refers to these as "dual-use" apps [5]. Determining how many users 
of a particular dual-use app are using it for intimate partner surveillance 
(IPS), and how many for its stated (usually legal) purpose, presents a 
significant challenge.

Therefore, we start by identifying a set of dual-use apps that appear to be 
marketed primarily for IPS. Apps that appear in Google searches for terms like 
"catch a cheating spouse," use vague language that appear to be marketed 
towards IPS, and label themselves as "spyware" became candidates that we 
considered. Such apps also differentiated themselves from more innocent apps by 
their price - they cost much more than a typical app, with monthly 
subscriptions ranging from \$12 to \$68 per month. We selected several of these 
to test. All of these apps brand themselves as spyware and advertise that their 
products are undetectable on the target phone, despite their disclaimers that 
they should not be used for any illegal activity. Some use language that 
appears designed to appeal to abusers. For example, underneath one promotional 
video, the app TheOneSpy has the disclaimer: "Note for this video: Don't 
confuse with the world "Loved Ones" as spouse or Girl friends! Word Loved ones 
has used for kids and teens." Many of these apps also refer to the tracked 
phone as the "target" phone, and advertise that once installed, the app is in 
"stealth mode," meaning completely undetectable.

The other category for stalkerware encompasses a large range of quality. Many 
apps advertised as stalkerware have absolutely no functionality other than 
advertisement of other apps. The few that are functional are often simplistic 
in both features and network patterns. Unlike dual-use apps, however, the vast 
majority are usually free with advertisements. These apps can be easily 
identified and found with a few keywords such as "spy","track", "locate", or 
"find".

@@

\section{Methodology}
\subsection{Finding Apps}
We selected thirty apps from a list provided by the team of Cornell researchers 
who performed the original large scale study of stalkerware\cite{chatterjee_spyware_2018}. An 
additional fifteen apps were included as they were recently added to the app 
store. Apps were chosen for their diversity of purpose and the probability that 
they can be used as stalkerware. These apps are the most likely to be 
downloaded when a abuser without deep technical knowledge tries to install 
stalkerware since they are either the first result upon searching for stalkerware, or are well known spyware apps due to advertisement or other reasons.

@@

%
% The next two lines define the bibliography style to be used, and the 
%bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{IPS_Measurement}

% 
% If your work has an appendix, this is the place to put it.
\appendix

>>>>

\section{Introduction}
In a world where technology now pervades every aspect of our lives, some 
aspects of our relationships with intimate partners have changed too. 
Specifically, when these relationships become abusive, there are more options 
available to the abuser for attempting to retain control over their partner. 
One of these options is the ability to install monitoring software on their 
partner's phone or computer to record their online activity, communications, 
and location. We refer to such software as ``stalkerware.''

Although stalkerware is superficially similar to other types of malware, it has 
a few key differences. First, stalkerware is meant to be installed 
deliberately, by someone with physical access to the device that is to be 
monitored. In the context of intimate partner abuse, this presents an enormous 
threat. An abuser will often know or be able to coerce passwords to the device 
from the victim, making traditional defense measures irrelevant. Second, the 
threat model used by malware developers includes sophisticated antivirus and 
intrusion detection systems (IDSs). Malware is therefore much more likely to 
attempt to obfuscate its network traffic. Third, stalkerware falls for the most 
part into two categories: apps that are explicitly advertised for surveilling 
spouses and partners, and apps that have a different stated use but can be 
repurposed to stalk a target. These are usually marketed for theft prevention, 
parental control of children, and employer control of employees. Previous 
literature refers to these as "dual-use" apps\cite{chatterjee_spyware_2018}. 
Some dual-use apps are ones that the survivor wants to keep installed, such as 
``find my phone'' apps or social media apps that leak the user's location. The 
most troubling example of this is Facebook, which most survivors want to keep 
so that they can contact their support network for help.  
Determining if a particular dual-use app is being used for 
intimate partner surveillance (IPS) presents an enormous challenge. Given the 
wide spectrum of apps that can be used to facilitate IPS and the limited scope 
of this project, we chose to focus on more overt spyware.

Largely due to the differences between stalkerware and other malware, most 
available information on the pervasiveness of stalkerware is anecdotal and 
piecemeal. Small-scale studies on how widespread spyware is do exist, but they 
have limitations. For example, in a survey of 72 
domestic violence shelters performed by NPR in 2014, 85\% said they worked with 
survivors whose abusers had tracked their location using their smartphone's 
GPS. 75\% said they are working with clients who have had their conversations 
monitored by hidden stalkerware apps.\cite{shahani_smartphones_nodate} However, 
this study does not report how many individual clients experienced being 
digitally stalked. Additionally, it and similar work relies on 
reports from domestic violence shelters and health workers, which does not give 
a complete picture of the situation. Little to no data exist regarding spyware 
use on IPA victims who never visit a shelter. Traditionally, antivirus vendors 
collect metrics about 
malicious software, but antivirus often does not detect spyware 
\cite{chatterjee_spyware_2018}. Alarmingly, the stalkerware 
websites themselves claim to have hundreds of thousands of downloads per app, 
but these numbers are not likely to be trustworthy. The goal of this work is 
therefore to identify a means of identifying stalkerware in the wild so that it 
can be enumerated. We collected the network signatures of 24 stalkerware apps 
and showed that they could be easily distinguished from ordinary smartphone 
network traffic. We also propose a simple method of generalizing our signatures 
to spyware that we did not manually examine, and evaluate its robustness.

We began by identifying a set of apps that appear to be 
marketed primarily for IPS. We focused exclusively on Android apps, since our 
two test phones were both Androids. Apps that appear in Google searches for 
terms like 
"catch a cheating spouse," use vague language that appear to be marketed 
towards IPS, and label themselves as "spyware" became candidates that we 
considered. Overt spyware is not generally available from the app store - it 
must be downloaded from the manufacturer sites. It also costs much more than a 
typical app, with monthly 
subscriptions ranging from \$12 to \$70 per month. The exception is an app that 
we call SpyToMobile, which survives on the app store by frequently 
``reskinning'' itself (see section \ref{reskinning}).  
We selected several of these to test. All of these apps brand themselves as 
spyware and advertise that their products are undetectable on the target phone, 
despite also having disclaimers in small text that their software should not be 
used for any illegal activity. Some use language that appears designed to 
appeal to abusers. For example, underneath one promotional 
video, the app TheOneSpy has the disclaimer: ``Note for this video: Don't 
confuse with the world `Loved Ones' as spouse or Girl friends! Word Loved ones 
has used for kids and teens.'' Many of these apps also refer to the tracked 
phone as the ``target'' phone, and advertise that once installed, the app is in 
``stealth mode,'' meaning completely undetectable. None of these required 
rooting the phone. When installed, they ask for all the permissions they 
require, including device administrator privileges. They are capable of 
blocking incoming calls, recording messages that have been deleted, recording 
the screen of the phone using accessibility features, monitoring apps such as 
Facebook, WhatsApp, Kik, and Viber, and more.

Other apps were overt spyware, but nevertheless available from the Play Store. 
They fell into two categories: spyware that reskinned itself to appear to be 
many different apps, and apps designed for parents to keep track of children 
(or in one case, for adult children to track elderly parents.) For the most 
part, this latter category was offered by antivirus companies such as Avast and 
Norton. These apps offered 
multiple features, such as call screening, blocking Google search results, 
recording text messages and call logs, and ``panic buttons'' that would allow 
the surveilled phone to contact the surveillant and raise an alarm. They do not 
hide their app icons and generally make an effort to notify the user that the 
monitoring app is installed on their phone. The spyware that did not notify 
users of its presence mostly consisted of one app, called SpyToMobile, that was 
provided on the Play Store under many different names.

We also identified many apps that were designed to track location or text 
messages, but were less comprehensive than the overt, off-store spyware. They 
track one feature each instead of tracking location, communication, and 
browsing history all at once. For the most part, these apps were free and 
available on the Google Play Store. Most of these were designed to keep track 
of the locations of the users' friends.

\section{The Stalkerware Ecosystem}
\subsection{The Practice of ``Reskinning''}
\label{reskinning}

Somewhere: put a note about the weird thing where the paid apps try to make you 
download stuff again. Maybe it's just us? Maybe they're trying to boost 
download count?

Some spyware has a notification that the app is running, but that can be 
disabled. Those ones don't hide their icon, but you might have to go to All 
Apps to find it.

\subsection{App Quality}

The other category for stalkerware encompasses a large range of quality. Many 
apps advertised as stalkerware have absolutely no functionality other than 
advertisement of other apps. The few that are functional are often simplistic 
in both features and network patterns. Unlike the overt apps, however, the vast 
majority are usually free with advertisements. These apps can be easily 
identified and found with a few keywords such as "spy","track", "locate", or 
"find".

@@

\section{Methodology}
\subsection{Finding Apps}
We selected thirty apps from a list provided by the team of Cornell researchers 
who performed the original large scale study of stalkerware[cite]. An 
additional fifteen apps were included as they were recently added to the app 
store. Apps were chosen for their diversity of purpose and the probability that 
they can be used as stalkerware. These apps are the most likely to be 
downloaded when a abuser without deep technical knowledge tries to install 
stalkerware.

@@

\section{Future Work}

Make machine learning identify stalkerware on the app store, download it, 
profile it. Then train more ML to pick out the malicious signatures? Would that 
even work? What would it learn that we can't?

\section{Conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{IPS_Measurement}

\appendix
